Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis.

### 2024-01-29 paper/paper.Rmd
testing website

As a simple example, we simulate two groups with $n = 100$ participants responding to an item ($Y$) with $k = 4$ ordered options. As explained in the previous sections, we can summarise the effect size of a CM (assuming PO) using a single $\beta = \log(OR) = \log(3)$. For simplicity, the probabilities of the first group $x = 0$ are uniform thus $P(Y = 1|x_0) = \dots P(Y = k|x_0) = 1/k$. The following code summarise the first steps of the simulation. Basically we define the simulation parameters, we calculate the $k - 1$ thresholds $\alpha$ and we apply Equations \@ref(eq:prob-cum-model1) and \@ref(eq:prob-cum-model2) calculating the probability of each response.

The other parameters to define are the thresholds. We can consider them as intercepts thus the transformed probabilities of each ordinal response when $x = 0$ (in this case the first group). The `alpha_to_prob()` and `prob_to_alpha()` can be used to go back and forth from thresholds and probabilities.

@@

lp <- lapply(alpha, function(a) a - log(or) * dat$x)
names(lp) <- sprintf("lp_leq%s", 1:(k - 1)) # giving appropriate names
lp <- data.frame(lp)
head(lp)

@@

# apply the inverse of the link function (invlogit) to calculate cumulative probabilities
cump <- lapply(lp, plogis)
cump <- data.frame(cump)
names(cump) <- sprintf("cump_leq%s", 1:(k - 1)) # giving appropriate names
head(cump)

@@

p <- data.frame(do.call(rbind, p)) # collapse list of rows into a dataframe
names(p) <- sprintf("p%s", 1:k) # giving appropriate names
head(p)

@@

# example of a random Y outcome based on the probabilities
sample(x = 1:k, size = 1, prob = p[1, ])

@@

dat$y <- apply(p, 1, function(ps) sample(1:k, size = 1, prob = ps))
head(dat)

@@

1.  Define simulation parameters, baseline probabilities (`prob0`), sample size, etc.
4.  Define regression coefficients $\boldsymbol{\beta}$ and calculate the $k - 1$ linear predictors ($\eta$) using the equations
5.  Apply the inverse of the link function $g^{-1}(\eta)$ on the linear predictor and calculate the cumulative probabilities $p(y \leq 1|x), p(y \leq 2|x), ... p(y \leq k - 1|x)$.
6.  Calculate the probabilities of $k$ outcomes
7.  Sample $n$ outcomes from a *multinomial* distribution, choosing between $k$ alternatives using the calculated probabilities
8.  Fit the appropriate model using `ordinal::clm()`

@@

A more efficient way to simulate an ordinal outcome is using the latent formulation of the model. This require simulating a standard linear regression using the appropriate data generation function (*logistic* or *normal*) and cutting the latent variable using the thresholds. The workflow is slightly different compared to the previous approach.

1.  Define simulation parameters as in the previous simulation. `prob0` are the probabilities when all predictors $\mathbf{X}$ are zero.
2.  Define regression coefficients $\boldsymbol{\beta}$ and calculate the linear predictor $\eta$ using the Equation \@ref(eq:latent-model)
3.  Add the random errors $\epsilon_i$ sampling from the chosen distribution (logistic or normal)
4.  Cut the latent variable into $k$ areas using the thresholds $\alpha$ and assign the corresponding ordinal value. This can be done using the `cut()` or the `findInterval()` functions.

@@

set.seed(2024)
N <- 1e3 # sample size
or <- 4 # odds ratio, higher here just for a more clear plot
k <- 4 # number of ordinal alternatives
probs0 <- rep(1/k, k) # probabilities for the first group
alpha <- prob_to_alpha(probs0, link = "logit")
dat <- data.frame(x = rep(c(0, 1), each = N/2))

@@

The simulation using the latent formulation of the model is implemented in the `sim_ord_latent()` function. Basically, we define the dataset `dat` with predictors. Then the model formula is specified within the function as `location = ~` along with the vector of regression coefficients (`beta`), baseline probabilities (`prob0`) and the link function. The function return a dataset with the simulated $Y$ and the latent variable (this is possible only because we simulate the data, otherwise $Y^{\star}$ cannot be observed by definition).

@@

For *probit* models we can set the $\beta_j$ to be in standardized (i.e., Cohen's $d$-like) units. For a categorical variable as the previous example with the group, $\beta_j$ is the degree of separation in standard deviation unit between the two latent distributions. For *logit* models we can set the odds ratio. Meaningful odds ratios can be derived from previous literature, meta-analyses or converting to other effect sizes. For example, @Sanchez-Meca2003-ji proposed some equations to convert between odds ratios and Cohen's $d$. Using their approach, a Cohen's $d = 0.5$ usually considered a plausible medium effect size corresponds to and odds ratio of $\approx 2.47$.

@@

```{r fig-example-cat-latent, fig.cap=filor::get_caption(), fig.place = TRUE}
cp <- cat_latent_plot(location = c(0, 1), scale = 1, prob0 = rep(1/4, 4), link = "logit", plot = FALSE) 
legend <- cowplot::get_legend(cp$probs_plot)
plt <- cowplot::plot_grid(
  cp$lat_plot + rm_legend(),
  cp$probs_plot + rm_legend()
)
cowplot::plot_grid(plt, legend, nrow = 2, rel_heights = c(0.95, 0.05))
```

@@

The parameters interpretation is the same as introduced in the general case. The thresholds $\alpha$ are fixed representing the probabilities $P(Y = k)$ when all predictors are zero. In this case by doing `alpha_to_prob(fit$alpha, link = "probit")` we should recover the `probs0` vector. `x1` is the main effect thus the difference in $z$ scores between *a* and *b* averaging over `x2`. The same holds for `x2`. `x1:x2` is the interaction thus difference of differences in $z$ scores (see Figure \@ref(fig:fig-effects-2-by-2-interaction)).

```{r fig-effects-2-by-2-interaction, fig.place = TRUE, fig.cap=filor::get_caption()}
# plot(ggpredict(fit, terms = c("x1", "x2")))
ggpredict(fit, terms = c("x1", "x2")) |> 
  data.frame() |>
  mutate(response.level = paste("Y = ", response.level)) |> 
  ggplot(aes(x = x, y = predicted, color = group)) +
  geom_point(size = 4) +
  geom_line(aes(group = group), linewidth = 1) +
  facet_grid(~response.level) +
  ylab("Probability") +
  viridis::scale_color_viridis(discrete = TRUE)
```

@@

Again, `x` is the slope of the numerical predictor averaging over `g` (given that `g` has been coded with sum-to-zero contrasts) thus the increase in $z$ scores for a unit increase in `x`. `g1` is the main effect of the factor evaluated when $x = 0$ (centering $x$ will change the parameter interpretation). The `x:g1` is the slopes difference the two groups. Figure \@ref(fig:fig-effects-num-by-cat-interaction) depicts the model results.

@@

@@

(ref:fig-show-th-example) Example of the `show_alpha()` function. On the left the assumed latent variable and $k - 1$ thresholds with the corresponding probabilities on the right.

@@

(ref:tab-model-summary-docx) Summary of the probit and logit models in terms of link ($g(\cdot)$) and inverse link function ($g^{-1}(\cdot)$) and the corresponding R code.

(ref:tab-model-summary-latex) Summary of the probit and logit models in terms of link ($g(\cdot)$) and inverse link function ($g^{-1}(\cdot)$) and the corresponding R code.

>>>>

As a simple example, we simulate two groups with $n = 100$ participants responding to an item ($Y$) with $k = 4$ ordered options. As explained in the previous sections, we can summarise the effect size of a CM (assuming PO) using a single $\beta = \log(OR) = \log(3)$. For simplicity, the probabilities of the first group $x = 0$ are uniform thus $P(Y = 1|x_0) = \dots P(Y = k|x_0) = 1/k$. The following code summarise the first steps of the simulation. We define the simulation parameters, calculate the $k - 1$ thresholds $\alpha$ and apply Equations \@ref(eq:prob-cum-model1) and \@ref(eq:prob-cum-model2) calculating the probability of each response.

To set the thresholds, we can consider them as intercepts thus the transformed probabilities of each ordinal response when $x = 0$ (in this case the first group). The `alpha_to_prob()` and `prob_to_alpha()` can be used to go back and forth from thresholds and probabilities.

@@

lp <- lapply(alpha, function(a) a - log(or) * dat$x)
names(lp) <- sprintf("lp_leq%s", 1:(k - 1)) # giving appropriate names
lp <- data.frame(lp)
head_tail(lp, n = 3)

@@

# apply the inverse of the link function (invlogit) to calculate cumulative probabilities
cump <- lapply(lp, plogis)
cump <- data.frame(cump)

# giving appropriate names, cump = cumulative probability, leq = less or equal
names(cump) <- sprintf("cump_leq%s", 1:(k - 1))
head_tail(cump, n = 3)

@@

p <- data.frame(do.call(rbind, p)) # collapse list of rows into a dataframe
names(p) <- sprintf("p%s", 1:k) # giving appropriate names

@@

# example of n random Y outcome based on the probabilities
sample(x = 1:k, size = 10, prob = p[1, ], replace = TRUE)

@@

dat$y <- apply(p, 1, function(ps) sample(1:k, size = 1, prob = ps))
head_tail(dat, n = 3)

@@

1.  Define simulation parameters, intercept probabilities (`prob0`), sample size, etc.
4.  Define regression coefficients $\boldsymbol{\beta}$ and calculate the $k - 1$ linear predictors ($\eta$) using the equations
5.  Apply the inverse of the link function $g^{-1}(\eta)$ on the linear predictor and calculate the cumulative probabilities $p(y \leq 1|x), p(y \leq 2|x), ... p(y \leq k - 1|x)$.
6.  Calculate the probabilities of $k$ outcomes
7.  Sample $n$ outcomes from a *multinomial* distribution using the calculated probabilities
8.  Fit the appropriate model using `ordinal::clm()`

@@

More efficiently, we can use the latent formulation of the CM. This require simulating a standard linear regression using the appropriate data generation function (*logistic* or *normal*) and cutting the latent variable using the thresholds. The workflow is slightly different compared to the previous approach.

1.  Define simulation parameters as in the previous simulation. `prob0` are the probabilities when all predictors $\mathbf{X}$ are zero.
2.  Define regression coefficients $\boldsymbol{\beta}$ and calculate the linear predictor $\eta$ using the Equation \@ref(eq:latent-model)
3.  Add the random errors $\epsilon_i$ sampling from the logistic or normal distributions
4.  Cut the latent variable into $k$ areas using the thresholds $\alpha$ and assign the corresponding ordinal value. This can be done using the `cut()` or the `findInterval()` functions.

@@

set.seed(2024)
N <- 1e3 # sample size
or <- 4 # odds ratio, higher here just for a more clear plot
k <- 4 # number of ordinal alternatives
probs0 <- rep(1/k, k) # probabilities for the first group
alpha <- prob_to_alpha(probs0, link = "logit") # thresholds
dat <- data.frame(x = rep(c(0, 1), each = N/2))

@@

The simulation using the latent formulation of the model is implemented in the `sim_ord_latent()` function. Basically, we define the dataset `dat` with predictors. Then the model formula is specified within the function as `location = ~` along with the vector of regression coefficients (`beta`), baseline probabilities (`prob0`) and the link function. The function return a dataset with the simulated $Y$ and the latent variable (this is possible only because we simulateds the data, otherwise $Y^{\star}$ cannot be observed by definition).

@@

For *probit* models we can set the $\beta_j$ to be in standardized (i.e., Cohen's $d$-like) units. For a categorical variable as the previous example with the group, $\beta_j$ is the degree of separation in standard deviation unit between the two latent distributions. For *logit* models we can set the odds ratio. Meaningful odds ratios can be derived from previous literature, meta-analyses or converting from other effect sizes. For example, @Sanchez-Meca2003-ji proposed some equations to convert between odds ratios and Cohen's $d$. Using their approach, a Cohen's $d = 0.5$ usually considered a plausible medium effect size corresponds to and odds ratio of $\approx 2.47$.

@@

```{r fig-example-cat-latent, fig.cap=filor::get_caption(), fig.place = TRUE}
cp <- cat_latent_plot(location = c(0, 1), scale = c(1, 1), prob0 = rep(1/4, 4), link = "logit", plot = FALSE)
legend <- cowplot::get_legend(cp$probs_plot)
plt <- cowplot::plot_grid(
  cp$lat_plot + rm_legend(),
  cp$probs_plot + rm_legend()
)
cowplot::plot_grid(plt, legend, nrow = 2, rel_heights = c(0.95, 0.05))
```

@@

The thresholds are fixed, representing the probabilities $P(Y = k)$ when all predictors are zero. In this case by doing `alpha_to_prob(fit$alpha, link = "probit")` we should recover the `probs0` vector. `x11` is the main effect thus the difference in $z$ scores between *a* and *b* averaging over `x2`. The same holds for `x21`. `x11:x21` is the interaction thus difference of differences in $z$ scores (see Figure \@ref(fig:fig-effects-2-by-2-interaction)).

```{r fig-effects-2-by-2-interaction, fig.place = TRUE, fig.cap=filor::get_caption()}
# plot(ggpredict(fit, terms = c ("x1", "x2")))
ggpredict(fit, terms = c("x1", "x2")) |> 
  data.frame() |>
  mutate(response.level = paste("Y = ", response.level)) |> 
  ggplot(aes(x = x, y = predicted, color = group)) +
  geom_point(size = 4) +
  geom_line(aes(group = group), linewidth = 1) +
  facet_grid(~response.level) +
  ylab("Probability") +
  viridis::scale_color_viridis(discrete = TRUE)
```

@@

Again, `x` is the slope of the numerical predictor averaging over `g` (given that `g` has been coded with sum-to-zero contrast) thus the increase in $z$ scores for a unit increase in `x`. `g1` is the main effect of the factor evaluated when $x = 0$ (centering $x$ will change the parameter interpretation). The `x:g1` is the difference between the slopes of the two groups. Figure \@ref(fig:fig-effects-num-by-cat-interaction) depicts the model results.

@@

# Conclusions

In this tutorial we presented the CM model in terms of model fitting, parameters interpretation and data simulation. The basic model can be expanded in several ways. When assuming PO or considering only a location effect is limiting, including scale (*location-scale*), thresholds (*location-shift*) effects or relaxing the PO assumption can help in modeling complex phenomenon. The supplementary materials contains some examples of these complex models. Furtermore, with clustered data (e.g., multiple items/repetitions for the same participant), a random-effect CM model can be fitted including random intercepts and slopes.

@@

(ref:fig-show-th-example) Example of the `show_alpha()` function. On the left the assumed latent variable and $k - 1$ thresholds with the corresponding probabilities on the right. The distance between thresholds (quantiles) determine the probability of the $k$ outcomes.

@@

(ref:tab-model-summary-docx) Summary of the probit and logit models in terms of link $g(\cdot)$ and inverse link function $g^{-1}(\cdot)$ and the corresponding R code.

(ref:tab-model-summary-latex) Summary of the probit and logit models in terms of link $g(\cdot)$ and inverse link function $g^{-1}(\cdot)$ and the corresponding R code.
